---
title: "Assessment 4"
author: "Nicole Hinojosa, Saqeena Thapa, Vanshika"
date: "2024-10-03"
output: html_document
---

# Part 1: Importing files, data wrangling, mathematical operations, plots and saving code on GitHub.


## Introduction
This report analyzes RNA-seq count data for gene expression and tree circumference measurements at two different sites over a 20-year period. 

## Task 1: RNA-seq count data for gene expression, high and low expression of 3 genes.

### 1.1 Read in the file “gene_expression.tsv”, making the gene identifiers the row names. Show a table of values for the first six genes.

1) Load libraries
```{r}
#RNA-seq Count Data Analysis
#Load necessary libraries
library(R.utils)
#use BiocManager::install("Biostrings") if it is not already installed in your Rstudio
library(Biostrings)
library(seqinr)
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
```

2) Read in the gene expression data
```{r}
#Download the data from the github link provided
URL = "https://raw.githubusercontent.com/ghazkha/Assessment4/refs/heads/main/gene_expression.tsv"
download.file(URL, destfile = "gene_expression.tsv")

# Read the downloaded TSV file into R
gene_expression <- read.table("gene_expression.tsv", header = TRUE, sep = "\t", row.names = 1)
```

3) 1st First 6 rows of the gene_expression data
```{r}
# View the first few rows of the data
head(n=6, gene_expression)
View( head (n=6, gene_expression))
```

### 1.2 Make a new column which is the mean of the other columns. Show a table of values for the first six genes.

Calculate Mean Expression
```{r}
# Calculate the mean across the samples and add as a new column
gene_expression <- gene_expression %>%
  mutate(mean_expression = rowMeans(select(., everything())))
# Show a table of values for the first six genes including the mean
head(n=6, gene_expression)
```

### 1.3 List the 10 genes with the highest mean expression.

Identify Top 10 Genes
```{r}
# List the 10 genes with the highest mean expression
top_genes <- gene_expression %>%
  arrange(desc(mean_expression)) %>%
  head(10)
# Print the top genes
print(top_genes)
```